---
title: "`simtrait`: Simulate Complex Traits from Genotypes"
author: "Alejandro Ochoa"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simtrait: Simulate Complex Traits from Genotypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, cache=FALSE, include=FALSE}
## copied from examples from the "simmer" R package
## after: https://www.enchufa2.es/archives/suggests-and-vignettes.html
## by Iñaki Úcar
required <- c("popkin", "bnpsd") # only suggested since simtrait doesn't need them to run...

if (!all(sapply(required, requireNamespace, quietly = TRUE)))
  knitr::opts_chunk$set(eval = FALSE)
```


In this vignette we first simulated an admixed population usinb `bnpsd`, then we simulate traits using `simtrait`.
In particular, we simulate a large number of traits to demonstrate that their sample covariance matrix is as expected.

## Load libraries required for this vignette

```{r}
library(popkin)   # to create plots of our covariance matrices
library(bnpsd)    # to simulate an admixed population
library(simtrait) # this package
```

## Simulate an admixed population

```{r}
# dimensions of data/model
m <- 10000 # number of loci
n_ind <- 30 # number of individuals, smaller than usual for easier visualizations
k <- 3 # number of intermediate subpops

# define population structure
F <- 1:k # FST values for k=3 subpopulations
s <- 0.5 # bias coeff of standard Fst estimator
Fst <- 0.3 # desired final Fst
obj <- q1d(n_ind, k, s=s, F=F, Fst=Fst) # data
# in this case return value is a named list with three items:
Q <- obj$Q # admixture proportions
F <- obj$F # rescaled Fst vector for intermediate subpops

# get pop structure parameters of the admixed individuals
Theta <- coanc(Q,F) # the coancestry matrix
Phi <- coanc_to_kinship(Theta) # kinship matrix

# draw allele freqs and genotypes
out <- rbnpsd(Q, F, m, wantP=FALSE, wantB=FALSE, noFixed=TRUE) # exclude variables not of interest
X <- out$X # genotypes
p_anc <- out$Pa # ancestral AFs
```

## Simulate a random trait

First we simulate one trait:

```{r}
# parameters of simulation
m_causal <- 100
herit <- 0.8

# create simulated trait
# case of exact p_anc
obj <- sim_trait(X=X, m_causal=m_causal, herit=herit, p_anc=p_anc)
# trait vector
length(obj$y)
n_ind
obj$y
# randomly-picked causal locus index
length( obj$i )
m_causal
head( obj$i ) # show partially...
# locus effect size vector
length( obj$beta )
m_causal
head( obj$beta ) # show partially...
```

## Compare sample covariance of trait to theoretical expectation

The interesting validation is simulation a large number of random traits, from which we can estimate a sample covariance matrix to compare to the desired theoretical one.

```{r}
# the theoretical covariance matrix of the trait is calculated by cov_trait
V <- cov_trait(kinship=Phi, herit=herit)

# simulate these many traits
n_traits <- 1000
# store in this matrix, initialize with zeroes
Y <- matrix(data=0, nrow=n_traits, ncol=n_ind)
# start loop
for (i in 1:n_traits) {
    obj <- sim_trait(X=X, m_causal=m_causal, herit=herit, p_anc=p_anc)
    Y[i,] <- obj$y # store in i^th row
}
# estimate sample covariance
V_sample <- cov(Y)
```

Now let's visualize the covariance matrices using `plotPopkin` from the `popkin` package.
Since both have large diagonals, we shrink them using `inbrDiag` also from the `popkin` package.

```{r, fig.width=6, fig.height=2.8, fig.align='center'}
# set outer margin for axis labels (left and right are non-zero)
par(oma=c(0,1.5,0,3))
# set inner margin for subpopulation labels (bottom and left are non-zero), add padding
# increase top margin for titles
par(mar=c(2,2,2,0)+0.2)
# now plot!
plotPopkin(
    list(inbrDiag(V), inbrDiag(V_sample)),
    titles=c('Theoretical', 'Sample Estimate'),
    legTitle='Covariance'
)
```
This plot verifies that the empirical covariance matches the theoretical expectation!

## Simulated trait without ancestral allele frequencies

For real data, true ancestral allele frequencies are unknown.
A reasonable, though imperfect, trait can still be simulated in these cases, but it is no longer possible to preserve the desired covariance structure, as shown below.
<!-- In particular, mean-centering the trait (done to set the desired mean) causes overfitting that not only reduces the overall covariance of the data, but introduces distortions to the covariance matrix. -->
The current implementation of `sim_trait` without `p_anc` relies on a known kinship matrix to introduce additional adjustments to the trait in order to ameliorate the discrepancy between the theoretical and the desired covariances.
However, the adjustments are not perfect.
<!-- However, some of the distortions remain. -->
A good kinship matrix estimate can be obtained using the `popkin` package.

For simplicity here we proceed using the true kinship matrix rather than an estimate:
```{r}
# store this in new matrix
Y2 <- matrix(data=0, nrow=n_traits, ncol=n_ind)
# start loop
for (i in 1:n_traits) {
    obj <- sim_trait(X=X, m_causal=m_causal, herit=herit, kinship=Phi)
    Y2[i,] <- obj$y # store in i^th row
}
# estimate sample covariance
V_sample2 <- cov(Y2)
```
Now we compare all three matrices:
```{r, fig.width=7, fig.height=2.8, fig.align='center'}
# set outer margin for axis labels (left and right are non-zero)
par(oma=c(0,1.5,0,3))
# set inner margin for subpopulation labels (bottom and left are non-zero), add padding
# increase top margin for titles
par(mar=c(2,2,2,0)+0.2)
# now plot!
plotPopkin(
    list(inbrDiag(V), inbrDiag(V_sample), inbrDiag(V_sample2)),
    titles=c('Theoretical', 'Sample Estimate 1', 'Sample Estimate 2'),
    legTitle='Covariance'
)
```

Here is a simple plot to compare scales, which shows 
<!-- the average agreement but also the distortions. -->
that the kinship is somewhat preserved but there is an excess overall covariance.
```{r}
plot(V, V_sample2)
abline(0,1, col='gray')
```
